---
output: github_document
bibliography: vignettes/dbscan.bib
link-citations: yes
---

```{r echo=FALSE, results = 'asis'}
pkg <- 'dbscan'

source("https://raw.githubusercontent.com/mhahsler/pkg_helpers/main/pkg_helpers.R")
pkg_title(pkg, anaconda = "r-dbscan", stackoverflow = "dbscan%2br")
```

## Introduction

This R package [@hahsler2019dbscan] provides a fast C++ (re)implementation of several density-based algorithms with a focus on the DBSCAN family for clustering spatial data.
The package includes: 
 
__Clustering__

- __DBSCAN:__ Density-based spatial clustering of applications with 
  noise [@ester1996density].
- __FOSC:__ Framework for optimal selection of clusters for unsupervised and 
  semisupervised clustering of hierarchical cluster trees [@campello2013density].
- __HDBSCAN:__  Hierarchical DBSCAN with simplified hierarchy 
  extraction [@campello2015hierarchical].
- __Jarvis-Patrick Clustering__: Clustering using a similarity measure based
  on shared near neighbors [@jarvis1973].
- __OPTICS/OPTICSXi:__ Ordering points to identify the clustering structure 
  and cluster extraction methods [@ankerst1999optics].
- __SNN Clustering__: Shared nearest neighbor clustering [@erdoz2003].


__Outlier Detection__

- __LOF:__ Local outlier factor algorithm [@breunig2000lof]. 
- __GLOSH:__ Global-Local Outlier Score from Hierarchies algorithm [@campello2015hierarchical]. 

__Cluster Evaluation__

- __DBCV:__ Density-based clustering validation [@moulavi2014].

__Fast Nearest-Neighbor Search (using kd-trees)__

- __kNN search__
- __Fixed-radius NN search__


The implementations use the kd-tree data structure (from library ANN) for faster k-nearest neighbor search, and are
for Euclidean distance typically faster than the native R implementations (e.g., dbscan in package `fpc`), or the 
implementations in [WEKA](https://ml.cms.waikato.ac.nz/weka/), [ELKI](https://elki-project.github.io/) and [Python's scikit-learn](https://scikit-learn.org/).

```{r echo=FALSE, results = 'asis'}
pkg_usage(pkg)
pkg_install(pkg)
```

## Usage

Load the package and use the numeric variables in the iris dataset
```{r}
library("dbscan")

data("iris")
x <- as.matrix(iris[, 1:4])
```

DBSCAN
```{r}
db <- dbscan(x, eps = .42, minPts = 5)
db
```

Visualize the resulting clustering (noise points are shown in black).
```{r dbscan}
pairs(x, col = db$cluster + 1L)
```


OPTICS
```{r}
opt <- optics(x, eps = 1, minPts = 4)
opt
```

Extract DBSCAN-like clustering from OPTICS 
and create a reachability plot (extracted DBSCAN clusters at eps_cl=.4 are colored)
```{r OPTICS_extractDBSCAN, fig.height=3}
opt <- extractDBSCAN(opt, eps_cl = .4)
plot(opt)
```

HDBSCAN

```{r}
hdb <- hdbscan(x, minPts = 4)
hdb
```

Visualize the hierarchical clustering as a simplified tree. HDBSCAN finds 2 stable clusters.

```{r hdbscan, fig.height=4}
plot(hdb, show_flat = TRUE)
```

## Additional guides

- [Using dbscan with tidyverse](https://michael.hahsler.net/dbscan/articles/tidyverse.html)
- [Using dbscan from Python](https://michael.hahsler.net/dbscan/articles/python.html)

## License 
The dbscan package is licensed under the [GNU General Public License (GPL) Version 3 or later](https://www.gnu.org/licenses/gpl-3.0.en.html). 

The __OPTICSXi__ R implementation in `R/optics_extractXi.R` was directly ported 
from the ELKI framework's Java implementation with permission by the original 
author, Erich Schubert. This function is redistributed under 
the stricter [GNU AGPLv3](https://www.gnu.org/licenses/agpl-3.0.en.html). 
Remove the file and function to use the package under the GNU GPL v3 license.  

## Citation request
```{r echo=FALSE, results = 'asis'}
pkg_citation(pkg, 2)
```
  
## References
