---
output: github_document
---

```{r echo=FALSE, results = 'asis'}
pkg <- 'dbscan'

source("https://raw.githubusercontent.com/mhahsler/pkg_helpers/main/pkg_helpers.R")
pkg_title(pkg)
```

This R package provides a fast C++ (re)implementation of several density-based algorithms with a focus on the DBSCAN family for clustering spatial data.
The package includes: 
 
__Clustering__

- __DBSCAN:__ Density-based spatial clustering of applications with noise.
- __HDBSCAN:__  Hierarchical DBSCAN with simplified hierarchy extraction.
- __OPTICS/OPTICSXi:__ Ordering points to identify the clustering structure clustering algorithms.
- __FOSC:__ Framework for Optimal Selection of Clusters for unsupervised and semisupervised clustering of hierarchical cluster tree.
- __Jarvis-Patrick clustering__
- __SNN Clustering__: Shared Nearest Neighbor Clustering.

__Outlier Detection__

- __LOF:__ Local outlier factor algorithm. 
- __GLOSH:__ Global-Local Outlier Score from Hierarchies algorithm. 

__Fast Nearest-Neighbor Search (using kd-trees)__

- __kNN search__
- __Fixed-radius NN search__

The implementations use the kd-tree data structure (from library ANN) for faster k-nearest neighbor search, and are typically faster than the native R implementations (e.g., dbscan in package `fpc`), or the 
implementations in [WEKA](https://www.cs.waikato.ac.nz/ml/weka/), [ELKI](https://elki-project.github.io/) and [Python's scikit-learn](https://scikit-learn.org/).

```{r echo=FALSE, results = 'asis'}
pkg_install(pkg)
```

## Usage

Load the package and use the numeric variables in the iris dataset
```{r}
library("dbscan")

data("iris")
x <- as.matrix(iris[, 1:4])
```

DBSCAN
```{r}
db <- dbscan(x, eps = .4, minPts = 4)
db
```

Visualize the resulting clustering (noise points are shown in black).
```{r dbscan}
pairs(x, col = db$cluster + 1L)
```


OPTICS
```{r}
opt <- optics(x, eps = 1, minPts = 4)
opt
```

Extract DBSCAN-like clustering from OPTICS 
and create a reachability plot (extracted DBSCAN clusters at eps_cl=.4 are colored)
```{r OPTICS_extractDBSCAN, fig.height=3}
opt <- extractDBSCAN(opt, eps_cl = .4)
plot(opt)
```

HDBSCAN

```{r}
hdb <- hdbscan(x, minPts = 4)
hdb
```

Visualize the hierarchical clustering as a simplified tree. HDBSCAN finds 2 stable clusters.

```{r hdbscan, fig.height=4}
plot(hdb, show_flat = TRUE)
```


## License 
The dbscan package is licensed under the [GNU General Public License (GPL) Version 3](https://www.gnu.org/licenses/gpl-3.0.en.html). The __OPTICSXi__ R implementation was directly ported from the ELKI framework's Java implementation (GNU AGPLv3), with explicit permission granted by the original author, Erich Schubert.  

## Changes
* List of changes from [NEWS.md](https://github.com/mhahsler/dbscan/blob/master/NEWS.md)

## References

- Hahsler M, Piekenbrock M, Doran D (2019). [dbscan: Fast Density-Based Clustering with R](https://doi.org/10.18637/jss.v091.i01). _Journal of Statistical Software_, *91*(1), 1-30. doi: 10.18637/jss.v091.i01.
- Martin Ester, Hans-Peter Kriegel, Joerg Sander, Xiaowei Xu (1996). A Density-Based Algorithm for Discovering Clusters in Large Spatial Databases with Noise. Institute for Computer Science, University of Munich. Proceedings of 2nd International Conference on Knowledge Discovery and Data Mining (KDD-96), 226-231. https://dl.acm.org/doi/10.5555/3001460.3001507
- Breunig, M., Kriegel, H., Ng, R., and Sander, J. (2000). LOF: identifying density-based local outliers. In ACM Int. Conf. on Management of Data, pages 93-104. doi: https://doi.org/10.1145/335191.335388
- Mihael Ankerst, Markus M. Breunig, Hans-Peter Kriegel, Joerg Sander (1999). OPTICS: Ordering Points To Identify the Clustering Structure. ACM SIGMOD international conference on Management of data. ACM Press. pp. doi: https://doi.org/10.1145/304181.304187
- Campello RJGB, Moulavi D, Zimek A, Sander J (2015). Hierarchical density estimates for data clustering, visualization, and outlier detection. ACM Transactions on Knowledge Discovery from Data (TKDD), 10(5):1-51. doi: https://doi.org/10.1145/2733381
- R. A. Jarvis and E. A. Patrick. 1973. Clustering Using a Similarity Measure Based on Shared Near Neighbors. IEEE Trans. Comput. 22, 11 (November 1973), 1025-1034. doi: https://doi.org/10.1109/T-C.1973.223640
- Levent Ertoz, Michael Steinbach, Vipin Kumar, Finding Clusters of Different Sizes, Shapes, and Densities in Noisy, High Dimensional Data, SIAM International Conference on Data Mining, 2003, 47-59. doi: https://doi.org/10.1137/1.9781611972733.5

