---
title: "Nearest-Neighbor Search and Graphs"
author: "Michael Hahsler"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Nearest-Neighbor Search and Graphs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
library(dbscan)
```

## Introduction

Nearest-neighbor search is the computational foundation for the clustering
and outlier-detection algorithms in **dbscan**. Many algorithms define 
neighborhoods using a `minPts` parameter. The neighborhood typically includes 
the point at the center as well, while `kNN` algorithms do not it in `k`.
Therefore, we have `k = minPts - 1` when we go from clustering to `kNN` search.

The package exposes `kNN`
search algorithms and resulting graphs directly:

| Function | Neighborhood | Result structure |
|:--|:--|:--|
| `kNN()` | The `k` closest observations | Matrices of neighbor IDs and distances |
| `frNN()` | All observations no farther than `eps` | Lists of neighbor IDs and distances |
| `sNN()` | Similarity based on shared nearest neighbors | Matrices of IDs, distances, and shared-neighbor counts |
| `kNNdist()` | Distance to the kth nearest neighbor | Numeric vector, or a distance matrix for several neighbors |
| `kNNdistplot()` | Sorted kth-neighbor distances | Diagnostic plot |

Objects returned by the first three functions inherit from class `NN`. They
support `print()`, `sort()`, `plot()`, `adjacencylist()`, and `comps()` for
finding connected components.

Nearest neighbor search is optimized for Euclidean distance where it used 
kd-tree implemented in the ANN library (Mount and Arya, 2010).

## Example data

We use the two-dimensional `moons` data included with the package. Row names
make it easier to follow neighbor IDs in the output.

```{r data}
data("moons")
x <- as.matrix(moons)
rownames(x) <- paste0("p", seq_len(nrow(x)))

plot(x, pch = 19, asp = 1, main = "Moons data")
```

For a numeric matrix or data frame, the functions use Euclidean distance.
Variable scales therefore determine which observations are considered close.
Standardize variables when their units are not comparable and equal weighting
is appropriate for the application.

## k nearest neighbors

`kNN()` finds exactly `k` neighbors for every observation. Self-matches are
removed when searching the rows of `x` against themselves.

```{r knn}
knn5 <- kNN(x, k = 5)
knn5

head(knn5$id)
head(knn5$dist)
```

Rows correspond to observations in `x`, and columns are ordered from nearest
to farthest when `sort = TRUE`, the default. IDs are row positions in `x`; row
names are retained as matrix row names. For example, the neighbors and
distances for observation 10 are:

```{r inspect-knn}
i <- 10
data.frame(
  id = knn5$id[i, ],
  row_name = rownames(x)[knn5$id[i, ]],
  distance = knn5$dist[i, ]
)
```

A `kNN` object represents a directed graph: observation `j` can be among the
nearest neighbors of `i` without `i` being among the nearest neighbors of
`j`. `adjacencylist()` returns the graph as a list of integer vectors.

```{r knn-adjacency}
knn_adj <- adjacencylist(knn5)
knn_adj[1:3]
```

The graph can be plotted for small, low-dimensional data. The first two data
columns are used as coordinates.

```{r plot-knn}
plot(knn5, x, pch = 19, main = "5-nearest-neighbor graph")
```

### Reuse and reduce a search

A stored search can be reduced without rebuilding the kd-tree. This is useful
when an analysis needs several values of `k`: calculate the largest required
neighborhood once and retain its leading columns.

```{r reduce-knn}
knn10 <- kNN(x, k = 10)
knn3 <- kNN(knn10, k = 3)

c(stored = knn10$k, reduced = knn3$k)
head(knn3$id)
```

The requested `k` cannot exceed the number stored in the input object. A new
search is required to enlarge it.

### Query new points

Use `query` to find neighbors in reference data `x` for a different set of
points. The output has one row per query point, and neighbor IDs still refer to
rows of `x`.

```{r knn-query}
query <- rbind(
  left = c(-0.5, 0.5),
  right = c(1.5, 0.5)
)

query_knn <- kNN(x, k = 4, query = query)
query_knn$id
query_knn$dist
```

Plotting a query result colors the reference observations selected for each
query. The query points are added as crosses.

```{r plot-query}
plot(query_knn, x, col = "grey70", main = "Neighbors of query points")
points(query, pch = 4, lwd = 2, cex = 1.3)
```

Self-matches are not automatically removed in query mode. Thus, if a query row
is identical to a reference row, that reference observation can be returned at
distance zero.

## Fixed-radius nearest neighbors

`frNN()` finds every observation within radius `eps`. Since neighborhood sizes
vary, IDs and distances are stored as parallel lists rather than matrices.
Self-matches are excluded.

```{r frnn}
fr <- frNN(x, eps = 0.25)
fr

fr$id[1:3]
fr$dist[1:3]
summary(lengths(fr$id))
```

The adjacency-list representation is already stored in `id`, so
`adjacencylist(fr)` returns that list. Empty integer vectors identify
observations with no other point inside the radius.

```{r frnn-adjacency}
identical(adjacencylist(fr), fr$id)
which(lengths(fr$id) == 0L)
```

Like a kNN graph, a fixed-radius graph can be plotted.

```{r plot-frnn}
plot(fr, x, pch = 19, main = "Fixed-radius graph (eps = 0.25)")
```

### Reduce a stored radius

A fixed-radius result can be filtered to a smaller radius without repeating
the search. Start with the largest radius that later analyses will need.

```{r reduce-frnn}
fr_wide <- frNN(x, eps = 0.30)
fr_small <- frNN(fr_wide, eps = 0.12)

c(wide_edges = sum(lengths(fr_wide$id)),
  small_edges = sum(lengths(fr_small$id)))
```

The new radius cannot exceed the radius stored in the object. Fixed-radius
queries use the same `query` interface as `kNN()`.

```{r frnn-query}
query_fr <- frNN(x, eps = 0.25, query = query)
data.frame(
  query = rownames(query),
  neighbors = lengths(query_fr$id)
)
```

## Shared nearest neighbors

`sNN()` starts with a k-nearest-neighbor search and counts neighborhood
overlap. Each observation is treated as belonging to its own neighborhood for
this calculation. The `shared` matrix aligns with `id`: `shared[i, j]` is the
similarity between observation `i` and the neighbor stored in `id[i, j]`.

```{r snn}
snn <- sNN(x, k = 10)
snn

head(snn$id, 3)
head(snn$shared, 3)
table(snn$shared)
```

By default, rows are sorted by decreasing shared-neighbor count. Distances and
IDs are rearranged with the counts, so the columns are no longer necessarily
ordered by Euclidean distance.

Set `kt` to retain only graph edges with at least that many shared neighbors.
Removed entries are represented by `NA` and are omitted by
`adjacencylist()`.

```{r threshold-snn}
snn5 <- sNN(snn, kt = 5)
snn5$id[1:3, ]
adjacencylist(snn5)[1:3]
```

With `jp = TRUE`, an edge receives a nonzero similarity only when the two
observations are in each other's nearest-neighbor lists. This mutual-neighbor
rule is used by Jarvis--Patrick clustering and by the package's SNN clustering
implementation.

```{r mutual-snn}
snn_mutual <- sNN(knn10, k = 10, jp = TRUE)
c(
  all_edges = sum(!is.na(snn$id)),
  mutual_edges = sum(snn_mutual$shared > 0)
)
```

Passing a precomputed `kNN` object, as above, avoids repeating the neighbor
search. Shared-neighbor clustering is covered separately in the
`vignette("nnclustering")` vignette.

## Neighbor-distance diagnostics

`kNNdist()` returns the distance from each observation to its kth nearest
neighbor in original row order.

```{r knndist}
d5 <- kNNdist(x, k = 5)
head(d5)
summary(d5)
```

Use `all = TRUE` to retain the distances to every neighbor from 1 through `k`.

```{r knndist-all}
d_all <- kNNdist(x, k = 5, all = TRUE)
head(d_all)
```

`kNNdistplot()` sorts these values and plots them. A sharp increase can suggest
a range for the DBSCAN radius: points before the increase have nearby
neighbors, while points in the upper tail are relatively isolated.

```{r knndistplot}
kNNdistplot(x, k = 5)
```

Several neighborhood sizes can be compared in one plot.

```{r knndistplot-multiple}
kNNdistplot(x, k = c(1, 5, 10))
legend("topleft", legend = c("k = 1", "k = 5", "k = 10"),
       col = 1:3, lty = 1, bty = "n")
```

For selecting `eps` for DBSCAN, `minPts` can be supplied instead. DBSCAN
counts the point itself, while `kNNdist()` excludes it, so
`kNNdistplot(x, minPts = 6)` uses `k = 5`.

```{r knndistplot-minpts}
kNNdistplot(x, minPts = 6)
```

The knee is a diagnostic, not an automatic parameter estimate. It may be
unclear for data containing groups with different densities.

## Connected components

`comps()` finds connected components in an `NN` graph. In a kNN graph,
`mutual = FALSE` treats a one-directional neighbor relation as sufficient for
a connection; `mutual = TRUE` requires both observations to list each other.

```{r components-knn}
comp_directed <- comps(knn3, mutual = FALSE)
comp_mutual <- comps(knn3, mutual = TRUE)

c(
  components_any_direction = length(unique(comp_directed)),
  components_mutual = length(unique(comp_mutual))
)
```

Fixed-radius graphs are symmetric, so no `mutual` argument is needed.

```{r components-frnn}
fr_comp <- comps(fr)
table(fr_comp)

plot(x, col = fr_comp, pch = 19, asp = 1,
     main = "Components of the fixed-radius graph")
```

Thresholded `sNN` graphs can be handled in the same way.

```{r components-snn}
snn_comp <- comps(snn5)
table(snn_comp)
```

`comps()` also accepts a `dist` object and a distance threshold. This is
equivalent to components in the corresponding fixed-radius graph.

```{r components-dist}
comp_dist <- comps(dist(x), eps = 0.25)
comp_fr <- comps(fr)

# Component numbers are arbitrary; compare which pairs share a component.
all(outer(comp_dist, comp_dist, `==`) ==
      outer(comp_fr, comp_fr, `==`))
```

## Sorting results

Sorting can be skipped during a search and applied later. This can save time
when an algorithm only needs the set of neighbors. For `frNN`, `sort()` orders
each list by distance; for `kNN`, it orders matrix rows by distance; and for
`sNN`, it orders rows by decreasing shared-neighbor count.

```{r sorting}
fr_unsorted <- frNN(x, eps = 0.25, sort = FALSE)
fr_sorted <- sort(fr_unsorted)

data.frame(
  id = fr_sorted$id[[1]],
  distance = fr_sorted$dist[[1]]
)
```

Setting `decreasing = TRUE` reverses the normal distance order for `kNN` and
`frNN`. The default for `sNN` is already decreasing similarity.

## Using non-Euclidean distances

Supplying a `dist` object allows `kNN()`, `frNN()`, and `sNN()` to use another
dissimilarity measure. However, this means that the efficient kd-tree 
implementation cannot be used.

```{r other-distance}
d_manhattan <- dist(x, method = "manhattan")

knn_manhattan <- kNN(d_manhattan, k = 5)
fr_manhattan <- frNN(d_manhattan, eps = 0.25)
snn_manhattan <- sNN(d_manhattan, k = 10)

c(
  knn_metric = knn_manhattan$metric,
  frnn_metric = fr_manhattan$metric,
  snn_metric = snn_manhattan$metric
)
```

A `dist` object stores all pairwise dissimilarities and therefore requires
quadratic memory. It also cannot be used with separate query points. Check that
the selected dissimilarity, transformations, and variable weights are
meaningful for the application.

## Search strategies and performance

For numeric data, the `search` argument offers three strategies:

* `"kdtree"`, the default, builds a kd-tree and is typically the best starting
  point for low- or moderate-dimensional data;
* `"linear"` checks the observations without building a tree; and
* `"dist"` first calculates all pairwise Euclidean distances.

The strategies should agree apart from the selection of tied observations.

```{r search-strategies}
knn_tree <- kNN(x, k = 5, search = "kdtree")
knn_linear <- kNN(x, k = 5, search = "linear")
knn_dist <- kNN(x, k = 5, search = "dist")

c(
  tree_vs_linear = isTRUE(all.equal(knn_tree$dist, knn_linear$dist)),
  tree_vs_dist = isTRUE(all.equal(knn_tree$dist, knn_dist$dist))
)
```

`bucketSize` and `splitRule` control construction of the kd-tree. Their
defaults are appropriate for most uses. Setting `approx` above zero enables
approximate search, which can improve speed at the cost of omitting some true
neighbors. Assess the effect on the downstream analysis before using it.

Kd-trees generally become less effective as dimensionality grows. Linear
search, an application-specific dimension reduction, or a different distance
representation may be preferable in high-dimensional settings.

## Important details

* Input data and query points must be numeric and finite for kd-tree and linear
  search.
* A search of `x` against itself excludes self-matches; query searches do not.
* `kNN()` returns exactly `k` observations. If multiple candidates tie at the
  kth distance, one is retained and the others are omitted.
* `frNN()` returns every observation within `eps`, so neighborhood sizes can
  differ and may be zero.
* IDs always refer to row positions in the reference data. Preserve a separate
  stable identifier if rows may later be reordered.

## Reference

Mount, D. M. and Arya, S. (2010). *ANN: A Library for Approximate Nearest
Neighbor Searching*. [ANN project page](https://www.cs.umd.edu/~mount/ANN/).
