---
title: "OPTICS: Ordering Points to Identify the Clustering Structure"
author: "Michael Hahsler, Matthew Piekenbrock"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{OPTICS: Ordering Points to Identify the Clustering Structure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
library(dbscan)
```

OPTICS (Ordering Points To Identify the Clustering Structure) is a
density-based ordering algorithm. Like DBSCAN, it describes dense regions in
terms of an epsilon neighborhood and a minimum number of points. Unlike
DBSCAN, its main result is not a single clustering. OPTICS orders observations
and records their reachability distances, revealing clustering structure over
a range of density levels.

This vignette shows how to create and interpret an OPTICS ordering and how to
extract flat or hierarchical clusters from it.

## Create an OPTICS ordering

We generate four compact groups for a small reproducible example.

```{r data}
set.seed(2)
n <- 400
x <- cbind(
  x = runif(4, 0, 1) + rnorm(n, sd = 0.1),
  y = runif(4, 0, 1) + rnorm(n, sd = 0.1)
)

plot(x, pch = 19, asp = 1, main = "Example data")
```

`optics()` accepts a numeric matrix or data frame. For numeric data it uses
Euclidean distance and a fast kd-tree search.

```{r fit}
opt <- optics(x, minPts = 10)
opt
```

The most important result components are:

* `order`: row numbers of the observations in OPTICS order;
* `reachdist`: each observation's reachability distance; and
* `coredist`: each observation's core distance.

The distance vectors use the original row order. For example, reachability
distances in OPTICS order are obtained with:

```{r inspect}
head(opt$order, 10)
head(opt$reachdist[opt$order], 10)
```

The first reachability distance is undefined because the first point has no
predecessor. It is stored as `Inf`.

## Read the reachability plot

Plotting an OPTICS object produces a reachability plot. Each vertical bar is
one observation, arranged in OPTICS order rather than input order.

```{r reachability}
plot(opt)
```

Low-reachability valleys represent dense groups. Peaks separate groups, and
deeper valleys represent denser structure. The plot is therefore best read as
a density landscape, not as a scatter plot or a sequence in the original
data.

The ordering can also be displayed in the original coordinate space. This is
mainly useful for understanding the algorithm; the reachability plot is
usually clearer for real data.

```{r order}
plot(x, col = "grey70", pch = 19, asp = 1, main = "OPTICS order")
lines(x[opt$order, ], col = "grey40")
points(x[opt$order[1], , drop = FALSE], pch = 19, col = 2)
```

## Choose `minPts` and `eps`

`minPts` controls the scale at which density is evaluated. It includes the
point itself, so `minPts = 10` uses the distance to the ninth other nearest
neighbor for the core-distance calculation. Larger values average over larger
neighborhoods and generally produce a smoother reachability plot; smaller
values reveal finer structure but are more sensitive to local variation.

In OPTICS, `eps` is the maximum neighborhood radius explored by the algorithm.
It is primarily a computational upper bound, not usually the density threshold
used to define the final clusters. The default `eps = Inf` lets the package
estimate a sufficient finite radius from the data and `minPts`.

For large data sets, supplying a smaller upper bound can reduce computation,
but it must not exclude structure of interest. Multiple dashed bars in the
reachability plot indicate undefined distances and usually mean that `eps` is
too small. The first dashed bar is expected.

As with other distance-based methods, variable scales matter. Standardize
variables when their units are not comparable and scaling is substantively
appropriate.

## Extract a DBSCAN-like clustering

`extractDBSCAN()` cuts the reachability plot at a global threshold `eps_cl`.
This does not rerun OPTICS, so several thresholds can be explored using the
same ordering. The extraction closely resembles DBSCAN with the same
`minPts` and epsilon, although OPTICS can leave some DBSCAN border points
unassigned.

```{r extract-dbscan}
opt_db <- extractDBSCAN(opt, eps_cl = 0.065)
opt_db
plot(opt_db)
```

The dashed horizontal line is the extraction threshold. Bars are colored by
cluster; cluster `0` is noise and appears black with the default palette.
Cluster assignments are stored in original data order.

```{r plot-dbscan}
table(opt_db$cluster)

plot(
  x,
  col = opt_db$cluster + 1L,
  pch = ifelse(opt_db$cluster == 0, 4, 19),
  asp = 1,
  main = "Clusters extracted at eps_cl = 0.065"
)
```

Changing the cut illustrates structure at a different density level:

```{r recut}
opt_db_wide <- extractDBSCAN(opt, eps_cl = 0.07)
c(
  clusters = max(opt_db_wide$cluster),
  noise = sum(opt_db_wide$cluster == 0)
)
```

Choose `eps_cl <= opt$eps`. A useful threshold crosses stable valleys below
the peaks that separate them. If a single global cut cannot represent groups
of differing density, use the Xi extraction described next.

## Extract clusters with the Xi method

`extractXi()` identifies relative steep-down and steep-up regions in the
reachability plot. The `xi` value is the minimum relative change in
reachability used to recognize a boundary. This can find nested clusters and
clusters at different density levels without selecting one global distance
threshold.

```{r extract-xi}
opt_xi <- extractXi(opt, xi = 0.05)
opt_xi
opt_xi$clusters_xi
plot(opt_xi)
```

Each row of `clusters_xi` gives the start and end positions of a cluster in
OPTICS order. These intervals can be nested. The colored line segments below
the reachability plot show the intervals, while `cluster` contains a convenient
non-nested labeling in original data order for plotting.

```{r plot-xi}
plot(
  x,
  col = opt_xi$cluster + 1L,
  pch = ifelse(opt_xi$cluster == 0, 4, 19),
  asp = 1,
  main = "Top-level Xi cluster labels"
)
```

Set `minimum = TRUE` to retain minimal clusters that do not contain other Xi
clusters. This is useful when a non-overlapping, fine-grained solution is
needed.

```{r minimum-xi}
opt_xi_min <- extractXi(opt, xi = 0.05, minimum = TRUE)
opt_xi_min
```

The predecessor correction is enabled by default and removes a known artifact
of the original Xi extraction procedure. It should normally remain enabled.

## Convert the ordering to a dendrogram

An OPTICS reachability structure can be represented as a dendrogram. This is
often convenient for smaller data sets or for code that works with standard R
dendrogram objects.

```{r dendrogram}
dend <- as.dendrogram(opt)
plot(dend, leaflab = "none", ylab = "Reachability distance")
```

Conversion requires a complete reachability structure. If OPTICS was run with
an `eps` that is too small, rerun it with a larger value or the default before
converting.

## Predict labels for new observations

Prediction is available only after `extractDBSCAN()`, because prediction needs
a flat clustering and its global `eps_cl` threshold. Supply the original
training data along with the new observations.

```{r predict}
new_points <- rbind(
  c(0.2, 0.3),
  c(0.8, 0.8),
  c(2.0, 2.0)
)

predict(opt_db, newdata = new_points, data = x)
```

A new observation receives the label of its nearest non-noise training point
within `eps_cl`; otherwise it is labeled `0`. Apply the same transformations
used for the training data to new observations.

## Use another distance measure

To use a distance other than Euclidean distance, calculate a `dist` object and
pass it to `optics()`. This uses the supplied pairwise distances rather than
the kd-tree search.

```{r distance, eval=FALSE}
d <- dist(x, method = "manhattan")
opt_manhattan <- optics(d, minPts = 10)
plot(opt_manhattan)
```

Prediction from a precomputed-distance analysis is not generally available,
because distances from new observations to the training data are not stored in
the `dist` object.

## References

Ankerst, M., Breunig, M. M., Kriegel, H.-P., and Sander, J. (1999).
OPTICS: Ordering Points To Identify the Clustering Structure. *Proceedings of
the 1999 ACM SIGMOD International Conference on Management of Data*, 49--60.
[doi:10.1145/304181.304187](https://doi.org/10.1145/304181.304187).

Schubert, E. and Gertz, M. (2018). Improving the Cluster Structure Extracted
from OPTICS Plots. *Lernen, Wissen, Daten, Analysen (LWDA 2018)*, 318--329.

Hahsler, M., Piekenbrock, M., and Doran, D. (2019). dbscan: Fast Density-Based
Clustering with R. *Journal of Statistical Software*, 91(1), 1--30.
[doi:10.18637/jss.v091.i01](https://doi.org/10.18637/jss.v091.i01).
